preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-15 Manuscript/manuscript.Rmd
Abstract and intro polishing

**Background** High-throughput chromosome conformation capture technology (Hi-C) has revealed extensive DNA looping and folding into discrete 3D domains. Among these are Topologically Associated Domains (TADs) and chromatin loops, known to play critical roles in cellular processes like gene regulation and cell differentiation. The relatively low resolution of Hi-C data (regions of several kilobases in size) prevents precise mapping of domain boundaries. TAD and loop boundaries have been characterized by the presence of CTCF and members of cohesin complex binding sites, typically mapped at higher precision. Several methods have been developed to leverage genome annotation  data for predicting TAD/loop boundaries; however, they overlooked key characteristics of the data, such as spatial associations between an annotation and a boundary. Furthermore, the number of regions containing boundaries is much smaller than the rest of the genome, creating a class imbalance problem.

@@

**Results** Using different feature engineering and resampling strategies, we found spatial relationships (distance) between boundaries and annotations have the most predictive power. Transcription factor binding sites outperformed other genome annotation types. Random under-sampling significantly improved model performance. Boundaries predicted by preciseTAD were more enriched for CTCF, RAD21, SMC3, and ZNF143 sites and more conserved across cell lines, highlighting their higher biological significance.

**Conclusion** Our results outline strategies for predictive modeling of 3D genomic domains using genome annotation data. The precise identification of TAD boundaries will improve our understanding of how genomic regulators shapes the 3D structure of the genome.

@@

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of important 3-dimensional (3D) spatial constructs within the human genome such as chromatin loops [@rao20143d], Topologically Associated Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At kilobase scale, gene promoters are often connected via distal regulatory elements, such as enhancers, through chromatin loops, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. Further, at the megabase scale, TADs represent regions on the linear genome, that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. Likewise, disruption of boundaries demarcating TADs and loops promotes cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Therefore, precisely identifying the location of TAD and chromatin loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

Several methods have been proposed to identify the genomic coordinates of TAD boundaries (reviewed in [@Zufferey:2018aa; @forcato2017comparison]), and chromatin loops [@rao20143d; @Ay:2014aa; @Heinz:2010aa; @salameh2019supervised].  However, a key limitation among most of these TAD- and loop callers, referred hereafter as domain callers, is that they are heavily reliant on Hi-C data *resolution*. Resolution refers to the sizes of genomic regions (bins) used to segment the linear genome [@li2019comparison]. Lower resolutions contribute to larger bin sizes, which lead to uncertainty in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TADs and chromatin loops [@fudenberg2016formation].

Another limitation among domain callers is that they disregard prior knowledge about functional genomic annotations associated with TAD and loop boundaries. The insulator binding protein, CTCF, and additional cofactors such as SMC3 and RAD21 have been identified as components of the loop extrusion model, whereby DNA is extruded through cohesin rings forming chromatin loops [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. Additionally, distinct patterns of histone modifications have also been shown to be present at boundaries, including H3K36 timethylation (H3K36me3) [@lieberman2009comprehensive;@dixon2012topological;@libbrecht2015joint]. These genomic annotations are obtained using chromatin immunoprecipitation followed by high-throughput DNA sequencing (ChIP-seq). The resolution of ChIP-seq experiments is typically on the order of tens to hundreds of bases [@mendoza2009genome], well below the resolution of Hi-C data (tens of kilobases; 750 bp is the highest resolution of Hi-C data to date [@bonev2017multiscale]). Therefore, leveraging precisely mapped genomic annotations in a supervised machine learning framework provides the possibility of more precisely predicting the locations of TAD (or loop) boundaries.

Here we present *preciseTAD*, an R package for implementing an optimally tuned machine learning framework for precise identification of domain boundaries using genome annotation data. Our method utilizes the random forest (RF) algorithm trained on chromatin segmentation (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS) data. We introduce a systematic pipeline for building optimal boundary prediction classifier. We found that spatial associations (distance) between boundaries and annotations perform best, transcription factor binding sites further improve prediction accuracy, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level (annotating each base and predicting its boundary probability), *preciseTAD* employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) on prediction probabilities to select genome annotation-guided boundary coordinates. This approach circumvents resolution restrictions of Hi-C data, allowing for the precise detection of biologically meaningful boundaries. We demonstrate that predictions made by *preciseTAD* are more enriched for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines. This improved precision in TAD and loop boundary location can provide insight into the role of genomic regulators on the 3D genome organization. The *preciseTAD* R package is freely available on GitHub (https://github.com/dozmorovlab/preciseTAD) and on Bioconductor (submitted).

>>>>

**Background** High-throughput chromosome conformation capture technology (Hi-C) has revealed extensive DNA looping and folding into discrete 3D domains. Among these are Topologically Associated Domains (TADs) and chromatin loops, known to play critical roles in cellular processes like gene regulation and cell differentiation. The relatively low resolution of Hi-C data (regions of several kilobases in size) prevents precise mapping of domain boundaries. TAD and loop boundaries have been characterized by the presence of CTCF and members of cohesin complex binding sites, and other genomic annotations. They are typically defined at high resolution (~100bp), suggesting they can inform the precise location of domain boundaries. Several methods have been developed to leverage genome annotation  data for predicting TAD/loop boundaries; however, they overlooked key characteristics of the data, such as spatial associations between an annotation and a boundary. Furthermore, the number of regions containing boundaries is much smaller than the rest of the genome, creating a class imbalance problem.

@@

**Results** Using different feature engineering and resampling strategies, we found spatial relationships (distance) between boundaries and annotations have the most predictive power. Transcription factor binding sites outperformed other genome annotation types. Random under-sampling significantly improved model performance. Boundaries predicted by preciseTAD were more enriched for CTCF, RAD21, SMC3, and ZNF143 sites and more conserved across cell lines, highlighting their higher biological significance. The model trained in one cell line perform well in predicting boundaries in another cell line.

**Conclusion** Our results outline strategies for predictive modeling of 3D genomic domains using genome annotation data. The precise identification of domain boundaries will improve our understanding of how genomic regulators are associated with the 3D structure of the genome.

@@

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of 3-dimensional (3D) spatial constructs within the human genome such as chromatin loops [@rao20143d], Topologically Associated Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At kilobase scale, gene promoters are often connected via distal regulatory elements, such as enhancers, through chromatin loops, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. Further, at the megabase scale, TADs represent regions on the linear genome, that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. Likewise, disruption of boundaries demarcating TADs and loops promotes cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Therefore, precisely identifying the location of TAD and chromatin loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

Several methods have been proposed to identify genomic coordinates of TADs (reviewed in [@Zufferey:2018aa; @forcato2017comparison]), and chromatin loops [@rao20143d; @Ay:2014aa; @Heinz:2010aa; @salameh2019supervised].  However, a key limitation among most of these TAD- and loop callers is that they are heavily reliant on Hi-C data *resolution*. Resolution refers to the sizes of genomic regions (bins) used to segment the linear genome [@li2019comparison]. Lower resolutions contribute to larger bin sizes, which lead to uncertainty in the location of boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TAD and loop boundaries [@fudenberg2016formation].

Another limitation among domain callers is that they disregard prior knowledge about functional genomic annotations associated with TAD and loop boundaries, referred hereafter as domain boundaries. The insulator binding protein, CTCF, and additional cofactors such as SMC3 and RAD21 have been identified as components of the loop extrusion model, whereby DNA is extruded through cohesin rings forming chromatin loops [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa]. Additionally, distinct patterns of histone modifications have also been shown to be present at boundaries, including H3K36me3 [@lieberman2009comprehensive;@dixon2012topological;@libbrecht2015joint]. These genomic annotations are obtained using chromatin immunoprecipitation followed by high-throughput DNA sequencing (ChIP-seq). The resolution of ChIP-seq experiments is typically on the order of tens to hundreds of bases [@mendoza2009genome], well below the resolution of Hi-C data (tens of kilobases; 750 bp is the highest resolution of Hi-C data to date [@bonev2017multiscale]). Therefore, leveraging precisely mapped genomic annotations in a supervised machine learning framework provides the possibility of more precisely predicting the locations of domain boundaries.

We present *preciseTAD*, an R package implementing an optimally tuned machine learning framework for precise identification of domain boundaries using genome annotation data. Our method utilizes the random forest (RF) algorithm trained on chromatin segmentation (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS) data. We introduce a systematic pipeline for building optimal boundary prediction classifier. We found that spatial associations (distance) between boundaries and annotations perform best, transcription factor binding sites further improve prediction accuracy, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level (annotating each base and predicting its boundary probability), *preciseTAD* employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) on prediction probabilities to select genome annotation-guided boundary coordinates. This approach circumvents resolution restrictions of Hi-C data, allowing for the precise detection of biologically meaningful boundaries. We demonstrate that predictions made by *preciseTAD* are more enriched for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines, and the model trained in one cell line can predict boundaries in another cell line. This improved precision in domain boundary location can provide insight into the association between genomic regulators and the 3D genome organization. The *preciseTAD* R package is freely available on GitHub (https://github.com/dozmorovlab/preciseTAD) and on Bioconductor (submitted).
